---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "Readme_files/"
)
options(width = 100)
devtools::install(quite = TRUE, upgrade = "always")
```

[![R-CMD-check](https://github.com/schalkdaniel/compboostSplines/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/schalkdaniel/compboostSplines/actions/workflows/R-CMD-check.yaml) [![codecov](https://codecov.io/gh/schalkdaniel/compboostSplines/branch/main/graph/badge.svg?token=KGR22VAOHI)](https://codecov.io/gh/schalkdaniel/compboostSplines) [![License: LGPL v3](https://img.shields.io/badge/License-LGPL_v3-blue.svg)](https://www.gnu.org/licenses/lgpl-3.0)

## C++ Spline Implementation of Compboost

Pure spline implementation of [compboost](https://compboost.org). This repository can be used for spline regression and non-parametric modelling of numerical features. The functionality contains the basis creation, the Demmler-Reinsch-Orthogonalization, as well as a method to subtract matrix bases.

**Feel free to extend the algorithms, improve performance, or use for your own projets.**

## Installation

#### Developer version:

```{r, eval=FALSE}
devtools::install_github("schalkdaniel/compboostSplines")
```

## Examples

<!--
- [Spline basis](#spline-basis)
- [Spline regression](#spline-regression)
- [Demmler-Reinsch-Orthogonalization](#demmler-reinsch-orthogonalization)
-->

### Spline basis

To create a spline basis call `createSplineBasis()` (for dense matrices) and `createSparseSplineBasis()` (for sparse matrices):
```{r}
library(compboostSplines)

nsim = 100

# Sample data:
x = sort(runif(nsim, 0, 10))
y = 2 * sin(x) + rnorm(nsim, 0, 0.5)

# Calculate knots of given x values:
knots = createKnots(values = x, n_knots = 20, degree = 3)

# Create basis using that knots:
basis = createSplineBasis(values = x, degree = 3, knots = knots)
str(basis)

# You can also create sparse matrices:
basis_sparse = createSparseSplineBasis(values = x, degree = 3, knots = knots)
str(basis_sparse)

# Check if row sums add up to 1:
rowSums(basis)
```

### Spline regression

With that, it is straight forward to use the least squares estimator for spline regression:

```{r}
# Custom function to estimate the least squares estimator
myEstimator = function(X, y, penmat = 0) {
  xtx = crossprod(X)
  xty = crossprod(X, y)
  L = chol(xtx + penmat)
  z = backsolve(L, xty, transpose = TRUE)
  return(as.vector(backsolve(L, z)))
}

# Polynomial regression using b-splines:
beta = myEstimator(basis, y)

# 20 knots may tend to overfit on the data, lets try p-splines with a penalty term of 4!
penalty = 4

# Get penalty matrix:
K = penaltyMat(ncol(basis), differences = 2)
K[1:6, 1:6]

# Get new estimator:
beta_pen = myEstimator(basis, y, penalty * K)

# Lets visualize the curves:

library(ggplot2)
library(ggthemes)

plot_df = data.frame(
  x = x,
  y = y
)
spline_df = data.frame(
  "Spline" = c(rep("B-Spline", nsim), rep("P-Spline", nsim)),
  "x" = rep(x, 2),
  "y" = c(basis %*% beta, basis %*% beta_pen)
)

ggplot() + geom_point(data = plot_df, mapping = aes(x = x, y = y)) +
  geom_line(data = spline_df, mapping = aes(x = x, y = y, color = Spline)) +
  theme_tufte() +
  scale_color_brewer(palette = "Set1")
```


### Demmler-Reinsch-Orthogonalization

In order to compare different models such as a linear model and additive model (using splines) we need to set the degrees of freedom equally. The Demmler-Reinsch-Orthogonalization can be used to translate given degrees of freedom to a penalty term:
```{r}
# We use the basis and penalty matrix from above and specify 2 and 4 degrees of freedom:
(penalty_df2 = demmlerReinsch(t(basis) %*% basis, K, 2))
(penalty_df4 = demmlerReinsch(t(basis) %*% basis, K, 4))

# This is now used for a new estimator:
beta_df2 = myEstimator(basis, y, penalty_df2 * K)
beta_df4 = myEstimator(basis, y, penalty_df4 * K)

plot_df = data.frame(
  x = x,
  y = y
)
types = c("B-Spline", "P-Spline", "P-Spline with 2 df", "P-Spline with 4 df")
spline_df = data.frame(
  "Spline" = rep(types, each = nsim),
  "x" = rep(x, length(types)),
  "y" = c(basis %*% beta, basis %*% beta_pen, basis %*% beta_df2, basis %*% beta_df4)
)

ggplot() + geom_point(data = plot_df, mapping = aes(x = x, y = y)) +
  geom_line(data = spline_df, mapping = aes(x = x, y = y, color = Spline)) +
  theme_tufte() +
  scale_color_brewer(palette = "Set1")
```

